---
title: "Exam1Findings"
format: html
---

## Interactive Visualizations Exploration

# Source 1
Chapter 13: Interactive Graphs (from Modern Data Visualization)
https://rkabacoff.github.io/datavis/Interactive.html?utm.com
Key findings / takeaways:
- plotly is a powerful interface: you can wrap a ggplot object in ggplotly() to automatically get interactivity (tooltips, zoom, pan, legend toggling).
- You can customize which fields show up in the tooltip by mapping extra variables to aesthetic labels.
- The interactivity features (zoom, pan, reset, download) are built in by default.
```{r}
library(ggplot2)
library(plotly)

# Using the built-in mpg dataset
p <- ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(size = 3) +
  labs(x = "Engine displacement (L)",
       y = "Highway mpg",
       color = "Car class") +
  theme_minimal()

# Turn the ggplot into an interactive plotly object
interactive_p <- ggplotly(p)

# Print / display it
interactive_p
```

# Source 2
How to Create Interactive Visualizations in R
https://www.kdnuggets.com/how-to-create-interactive-visualizations-in-r?utm.com
Key findings / takeaways: 
- Packages to know: plotly for converting ggplots to interactive, or building directly with plot_ly(). leaflet for maps and spatial interactivity (zooming, panning maps). DT for interactive tables (sorting, filtering, pagination).
- The article emphasizes that interactive visualizations allow users to explore — zooming, filtering, and digging more deeply than a static plot allows.
- It also warns: interactivity isn’t always beneficial. If overdone, it can confuse users or overwhelm them.
```{r}
library(leaflet)
library(DT)

# Example: interactive map with leaflet
m <- leaflet() %>%
  addTiles() %>%
  addMarkers(lng = c(-71, -74), lat = c(42, 40),
             popup = c("Point A", "Point B"))
m

# Example: interactive table with DT
library(datasets)
df <- head(iris, 20)
datatable(df, options = list(pageLength = 5, filter = "top"))
```

# Source 3
The Impact of Interactive Data Visualization on Decision-Making in Business Intelligence
https://www.researchgate.net/publication/381265855_The_Impact_of_Interactive_Data_Visualization_on_Decision-Making_in_Business_Intelligence
Key findings / takeaways:
- Using interactive visualizations improves data exploration, speeds decision-making, and enhances collaboration among stakeholders.
- However, the benefit depends on data quality, user skill level, and interface design. Poor design or messy data can negate the advantages.
- The authors caution against overly complex interactions, which can cause cognitive overload in users and reduce effectiveness. ResearchGate



